Spatial Reasoning and Statistical Graphics

### 2014-10-22 Paper/Draft1/TestingVisualAptitude.Rnw
Move lit review list to the back
Add discuussion of spatial reasoning and statistical graphics.

\section{Introduction}
Relevant literature: 
\item \citet{shah1995conceptual} showed that spatial ability was not correlated with accuracy on a simple two-dimensional line graph description task, but that mathematical ability was correlated with accuracy. 

Lineups depend on the ability to search for a signal amid distractors (Visual Search Task) and the ability to infer patterns from stimuli (Pattern Recognition task). Some lineups (polar coords) also depend on the ability to mentally rotate stimuli (spatial rotation task) and mentally manipulate graphs (paper folding task). By breaking the lineup task down into component parts, we can correlate lineup performance with similar cognitive factor tests to determine where additional variation in skill level factors into performance differences. In addition, we can correlate previous experiences (science-based major, research experience, Auto-CAD skills) with performance to explore the effect that participant experience has on lineup performance.

\section{Methods}
\comment{Besides pointing out what we asked participants to do, we also need some why, and also why in particular these tests were selected. part of the validity consideration can go in here - and the classification from the naval test of the tests we picked into different classes (maybe already with a hint that for our data these weren't actually orthogonal classes, because we found a really high correlation between card rotation and paper folding.)\\
\textit{I don't know if the tests are supposed to be orthogonal. It would be hard to construct a set of tests that was orthogonal, because the abilities themselves aren't easily separable - math reasoning and spatial reasoning are related, for instance, and verbal abilities are correlated with creativity too.}}
Participants will complete the following tasks (sample pictures included, full stimuli set will be added to the appendix once we are sure there is no need for follow-up experiments). Tasks are designed so that participants are under time pressure; they are not expected to complete all of the problems in each section. This provides more discrimination between high scorers and prevents score compression at the top of the range. 
\begin{itemize}
\item Visual Search Task: designed to test participants' ability to find a target stimulus in a field of distractors. An example is shown in figure \ref{fig:VST}. The visual search task is similar in concept to lineups: it tests one's ability to find the target plot. Historically, it has been used as a measure of brain damage\citep{goldstein1973validity,demita1981validity,moerland1986neuropsychological}; however, similar tasks, have been used to measure cognitive performance in a variety of situations (under the influence of drugs, for example, in \citet{anderson1983interactive}). The similarity to lineup protocol as well as the simiplicity of the test and its' lack of color justify the slight deviation from forms of visual search tasks typically used in normal populations. 
\begin{figure}[htbp]\centering
\includegraphics[width=.7\textwidth]{VisualSearch}
\caption[Visual Search Task]{Visual Search Task. Participants are instructed to find the plot numbered 1-24 which matches the plot labeled "Target". Participants will complete up to 25 of these tasks in 5 minutes.}\label{fig:VST}
\end{figure}
\item Paper Folding Task: tests participants' ability to visualize and mentally manipulate figures in three dimensions. Associated with the ability to extrapolate symmetry and reflection over multiple steps. An example is shown in figure \ref{fig:paperfolding}. 
\begin{figure}[htbp]\centering
\includegraphics[width=.9\textwidth]{paperfolding}
\caption[Paper Folding Task]{Paper Folding Task. Participants are instructed to pick the figure matching the sequence of steps shown in the left-hand figure. Participants will complete up to 20 of these tasks in 6 minutes.}\label{fig:paperfolding}
\end{figure}
Lineups require similar manipulations in two-dimensional space, and also require the ability to perform complex spatial manipulations mentally (for instance, comparing the interquartile range of two boxplots as well as their relative alignment to a similar set of two boxplots in another panel).

@@

>>>>

\section{Introduction}
\comment{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs: \citet{shah1995conceptual} showed that mathematical ability, but not spatial ability, was associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: \citet{mayer1994} hypothesize that those with low spatial ability perform worse on tests utilizing diagrams and graphs because more cognitive resources are required to process the visual stimuli (leaving fewer resources to make connections and draw conclusions from those stimuli). Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner \citep{vekiri2002value}. Both of these theories suggest spatial ability would impact a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. Previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs \citep{lowrie2007solving}.

\subsection{The Lineup Protocol}
\comment{More lineup introduction here}

\newtext{Statistical lineups \citep{hofmann2012graphical,buja2009statistical} depend on the ability to search for a signal amid distractors (visual search) and the ability to infer patterns from stimuli (pattern recognition). Some lineups (polar coords) also depend on the ability to mentally rotate stimuli (spatial rotation) and mentally manipulate graphs (spatial rotation and manipulation). By breaking the lineup task down into component parts, we may be able to determine which visuospatial factors most strongly correlate with lineup performance, using carefully chosen cognitive tests to assess these aspects of visuospatial ability. These assessments will then be used to account for some of the variation in lineup performance differences. In addition, we will also examine the effect that previous experiences (science-based major, research experience, Auto-CAD skills) have on lineup performance and the implications for the use of lineups to assess the graphical display of statistical information. }

\section{Methods}
\comment{Use the setup in the lineup protocol section to justify each of the tests chosen. }
\comment{Besides pointing out what we asked participants to do, we also need some why, and also why in particular these tests were selected. part of the validity consideration can go in here - and the classification from the naval test of the tests we picked into different classes (maybe already with a hint that for our data these weren't actually orthogonal classes, because we found a really high correlation between card rotation and paper folding.)\\
\textit{I don't know if the tests are supposed to be orthogonal. It would be hard to construct a set of tests that was orthogonal, because the abilities themselves aren't easily separable - math reasoning and spatial reasoning are related, for instance, and verbal abilities are correlated with creativity too.}}
Participants will complete the following tasks (sample pictures included, full stimuli set will be added to the appendix once we are sure there is no need for follow-up experiments). Tasks are designed so that participants are under time pressure; they are not expected to complete all of the problems in each section. This provides more discrimination between high scorers and prevents score compression at the top of the range. 
\begin{itemize}
\item Visual Search Task: designed to test participants' ability to find a target stimulus in a field of distractors. An example is shown in figure \ref{fig:VST}. The visual search task is similar in concept to lineups: it tests one's ability to find the target plot. Historically, it has been used as a measure of brain damage\citep{goldstein1973validity,demita1981validity,moerland1986neuropsychological}; however, similar tasks, have been used to measure cognitive performance in a variety of situations (under the influence of drugs, for example, in \citet{anderson1983interactive}). The similarity to lineup protocol as well as the simiplicity of the test and its' lack of color justify the slight deviation from forms of visual search tasks typically used in normal populations. 
\begin{figure}[htbp]\centering
\includegraphics[width=.7\textwidth]{VisualSearch}
\caption[Visual Search Task]{Visual Search Task. Participants are instructed to find the plot numbered 1-24 which matches the plot labeled "Target". Participants will complete up to 25 of these tasks in 5 minutes.}\label{fig:VST}
\end{figure}
\item Paper Folding Task: tests participants' ability to visualize and mentally manipulate figures in three dimensions. Associated with the ability to extrapolate symmetry and reflection over multiple steps. An example is shown in figure \ref{fig:paperfolding}. 
\begin{figure}[htbp]\centering
\includegraphics[width=.9\textwidth]{paperfolding}
\caption[Paper Folding Task]{Paper Folding Task. Participants are instructed to pick the figure matching the sequence of steps shown in the left-hand figure. Participants will complete up to 20 of these tasks in 6 minutes.}\label{fig:paperfolding}
\end{figure}
Lineups require similar manipulations in two-dimensional space, and also require the ability to perform complex spatial manipulations mentally (for instance, comparing the interquartile range of two boxplots as well as their relative alignment to a similar set of two boxplots in another panel).

@@

\subsection{Lit Review Rejects}
